An efficient composition of bidirectional programs by tupling and lazy updates

### 2019-11-25 paper/2.minbigul.tex
update tex 20191125
- def of get_mh
- v / v'
- experiments: xpg is faster than pgs
- dummy
- kpg: lazy eval on fi-cond
- figure size
- typos

For source/view rearrangement, BiGUL use just one lambda expression to express how to deconstruct as well as reconstruct data. It is a kind of bijection. However, to be able to implement it in OCaml, the environment used for developing minBiGUL and solutions in the paper, we need give two functions which one is the inverse of the other. In the above syntax, $f_2 = f_1^{-1}$ and $g_2 = g_1^{-1}$.

@@

In definitions 1 and 2, we use if-then-else-fi statements to express semantics of $\putbxinline{Case}$ and $\getbxinline{Case}$. This statement is useful to describe many functions related to $Case$ in this paper. Statement (if $E_1$ then $X_1$ else $X_2$ fi $E_2$) means if the test $E_1$ is true, the statement $X_1$ is executed and the assertion $E_2$ must be true, otherwise, i.e. $E_2$ is false, the statement $X_2$ is executed and the assertion $E_2$ must be false. If the values of $E_1$ and $E_2$ are distinct, the if-then-else-fi structure is undefined. We can write the equivalent if-then-else statement as follows:

>>>>

For source/view rearrangement, BiGUL uses just one lambda expression to express how to deconstruct as well as reconstruct data. It is a kind of bijection. However, to be able to implement it in OCaml, the environment used for developing minBiGUL and solutions in the paper, we need give two functions which one is the inverse of the other. In the above syntax, $f_2 = f_1^{-1}$ and $g_2 = g_1^{-1}$.

@@

We define the semantics of $put$ and $get$ as in definitions 1 and 2 respectively. Instead of using $v'$ in the $put$ direction like figures \ref{fig:eval-phead}, \ref{fig:eval-comp-phead} and \ref{fig:eval-comp-phead-2}, we just simply use notation $v$ as the updated view. The later definitions also use this convention.

In the above definitions, we use if-then-else-fi statements to express semantics of $\putbxinline{Case}$ and $\getbxinline{Case}$. This statement is useful to describe many functions related to $Case$ in this paper. Statement (if $E_1$ then $X_1$ else $X_2$ fi $E_2$) means if the test $E_1$ is true, the statement $X_1$ is executed and the assertion $E_2$ must be true, otherwise, i.e. $E_2$ is false, the statement $X_2$ is executed and the assertion $E_2$ must be false. If the values of $E_1$ and $E_2$ are distinct, the if-then-else-fi structure is undefined. We can write the equivalent if-then-else statement as follows:

### 2019-11-25 paper/3.minbigul_m.tex
update tex 20191125
- def of get_mh
- v / v'
- experiments: xpg is faster than pgs
- dummy
- kpg: lazy eval on fi-cond
- figure size
- typos

For minBiGUL or even BiGUL, when evaluating a composition of many programs, $get$s are reevaluated so many times since no intermediate state is stored during the evaluation. To avoid reevaluating $get$s, we can save intermediate states in a key-value table where the key is a pair of the program $bx$ and the source $s$, and the value is the result of $\getbx{bx}{s}$. Then, searching an evaluated $get$s on the table is used instead of recomputing it. We define a memory version, minBiGUL$_m$, including $get_m$ and $put_m$ as in definition 3 and 4 respectively. Another function introduced in definition 5, $get_{mh}$, is used to retrieve the value of an evaluated $get$ from the table or compute an unevaluated $get$ by using $get_m$ and store the key-value tuple to the table.

\begin{definition} $\putm{bx}{s}{v}$

    $\putm{bx}{s}{v} = \text{same with } put \text{ if } bx \neq bx_1 \circ bx_2$

    $\putm{bx_1 \circ bx_2}{s}{v} = \putm{bx_1}{s}{(\putm{bx_2}{(\getmh{bx_1}{s})}{v})}$
\end{definition}

\begin{multicols}{2}
    \begin{definition} $\getm{bx}{s}{v}$

        $\getm{bx}{s} = \\
            \tab \text{same with } get \text{ if } bx \neq bx_1 \circ bx_2$

        $\getm{bx_1 \circ bx_2}{s} =\\
            \tab \text{try } (\text{Hashtbl.find} \ table_g \ (bx, s))\\
            \tab \text{with Not\_found } \to\\
                \qtab v \Leftarrow \getmh{bx_2}{(\getmh{bx_1}{s})}\\
                \qtab \text{Hashtbl.add } \ table_g \ (bx, s) \ v\\
                \qtab v$
    \end{definition}
\columnbreak
    \begin{definition} $\getmh{bx}{s}{v}$

        $\getmh{bx}{s} =\\
            \tab \text{try } (\text{Hashtbl.find} \ table_g \ (bx, s))\\
            \tab \text{with Not\_found } \to\\
                \qtab v \Leftarrow \getm{bx}{s}\\
                \qtab \text{try } (\text{Hashtbl.find} \ table_g \ (bx, s))\\
                \qtab \text{with Not\_found } \to\\
                    \qtabs{2} \text{Hashtbl.add } \ table_g \ (bx, s) \ v\\
                    \qtabs{2} v$
    \end{definition}
\end{multicols}

>>>>

For minBiGUL or even BiGUL, when evaluating a composition of many programs, $get$s are reevaluated so many times since no intermediate state is stored during the evaluation. To avoid reevaluating $get$s, we can save intermediate states in a key-value table where the key is a pair of the program $bx$ and the source $s$, and the value is the result of $\getbx{bx}{s}$. Then, searching an evaluated $get$s on the table is used instead of recomputing it. We define a memory version, minBiGUL$_m$, including $get_m$ and $put_m$ as in definition 3 and 4 respectively. Another function introduced in definition 5, $get_{mh}$, is used to retrieve the value of an evaluated $get$ from the table or compute an unevaluated $get$ by using $get_m$ then store the key-value tuple to the table if necessary.

\begin{definition} $\text{Memo version of } put$

    $\putm{bx}{s}{v} = \text{match } bx \text{ with }\\
        \tab \vert \ bx_1 \circ bx_2 \to \putm{bx_1}{s}{(\putm{bx_2}{(\getmh{bx_1}{s})}{v})} \\
        \tab \vert \ \_ \to \text{similar to } put$
\end{definition}

\begin{multicols}{2}
    \begin{definition} $\text{Memo version of } get$

        $\getm{bx}{s} = \text{match } bx \text{ with }\\
            \tab \ \vert \ bx_1 \circ bx_2 \to\\
            \tabs{2} \text{try } (\text{Hashtbl.find} \ table_g \ (bx, s))\\
            \tabs{2} \text{with Not\_found } \to\\
                \tabs{3} v \Leftarrow \getmh{bx_2}{(\getmh{bx_1}{s})}\\
                \tabs{3} \text{Hashtbl.add } \ table_g \ (bx, s) \ v\\
                \tabs{3} v \\
            \tab \ \vert \ \_ \to \text{similar to } get$
    \end{definition}
\columnbreak
    \begin{definition} $???$

        $\getmh{bx}{s} =\\
            \tab \text{try } (\text{Hashtbl.find} \ table_g \ (bx, s))\\
            \tab \text{with Not\_found } \to\\
                \tabs{2} v \Leftarrow \getm{bx}{s}\\
                \tabs{2} \text{match } bx \text{ with } \\
                \tabs{2} \ \vert \ bx_1 \circ bx_2 \to v \\
                \tabs{2} \ \vert \ \_ \to \text{Hashtbl.add } \ table_g \ (bx, s) \ v\\
                \qtabs{2} \ v$
    \end{definition}
\end{multicols}

### 2019-11-25 paper/4.xpg.tex
update tex 20191125
- def of get_mh
- v / v'
- experiments: xpg is faster than pgs
- dummy
- kpg: lazy eval on fi-cond
- figure size
- typos

% $pg$ is an involution that is self-inverse. An involution is a function $f$ that satifies $f(f(x)) = x \text{ for all } x \text{ in the domain of } f$.

@@

$\tab \pg{Skip \ h}{s}{v} \overset{1}{=} (\sif{h \ s = v}{s}{\text{fail}}, h \ s) \\
    \qtab \overset{2}{=} \sif{h \ s = v}{(s, h \ s)}{\text{fail}} \\
    \qtab \overset{3}{=} \sif{h \ s = v}{(s, v)}{\text{fail}}$\\
The first equality is simply based on the definitions of $pg$, $\putbxinline{Skip \ h}$ and $\getbxinline{Skip \ h}$. The second one tuples two results of $put$ and $get$ in the body of the if-expression. This is a trick since in some cases, the result of $pg$ may be fail although there is no fail when evaluating $\getbxinline{Skip \ h}$. The last equality is relatively obivious.

@@

A restriction for $\pginline{Case}$ needs to be introduced here. We know that there is one entering condition and one exit condition when evaluating $\putbxinline{Case}$ as well as $\getbxinline{Case}$. If a tupling occurs, there will be 4 combinations from these conditions. This means two entering conditions of $\putbxinline{Case}$ and $\getbxinline{Case}$ are not always simultaneously satified. The evaluated branches are distinct in the $put$ and $get$ directions for combinations $((cond_{sv} \ s \ v) \ \&\& \ (not (cond_{s} \ s)))$ and $((not (cond_{sv} \ s \ v) \ \&\& \ (cond_{s} \ s))$, which are restricted in this paper. This does not happen for the others which is used in the construction of $\pginline{Case}$.

@@

The construction of $\pginline{bx_1 \circ bx_2}$ can be considered as the soul of the $pg$ function. The first two equalities comes from mentioned definitions and some basic transformations. The third one rewrites $v_1 \Leftarrow \getbx{bx_1}{s}$ into $(s_1, v_1) \Leftarrow \pg{bx_1}{s}{dummy}$. It is possible if we consider $\getbx{bx_1}{s}$ as the second element of $\pg{bx_1}{s}{dummy}$ where $dummy$ is a special value that makes the $put \, [\![bx_1]\!]$ valid. Due to \textsc{GetPut} law, we can use $get \ s$ to update the source $s$ in the $put$ direction. Then it is feasible to construct a $dummy$ from the source. The last equality changes $dummy$ by an application $construct\_dummy \ s$, and also substitutes $s$ in $(s_3, v_3) \Leftarrow \pg{bx_1}{s}{s_2}$ by $s_1$ which is the evaluated result of $\putbx{bx_1}{s}{(construct\_dummy \ s)}$. This transformation is possible because of the \textsc{PutPut} law. Then, $v_3$ in the result pair $(s_3,v_3)$ equals $dummy$. So we can realize that there is no loss information when computing a composition.

@@

When evaluating $\pginline{bx_1 \circ bx_2}$, there are three $pg$ calls, of which twice for $\pginline{bx_1}$ and once for $\pginline{bx_2}$. If a given program is a left-associative composition, the number of $pg$ calls will be exponential. Therefore, the runtime inefficiency is inevitable. To solve that, we introduce a new function, $cpg$, accumulates changes in the source and the view. $\cpg{bx}{ks}{kv}{s}{v}$ is an extension of $\pg{bx}{s}{v}$ where $ks$ and $kv$ are continuations used to hold the modification information, and $s$ and $v$ are used to keep evaluated values. The output of this function is a 4-tuple $(ks, kv, s, v)$. To be more convenient for presenting the definition of $cpg$ as well as the other functions later, we provide some following utility functions:\\ $fst = \lambda (x_1,x_2). x_1 , \quad snd = \lambda (x_1,x_2). x_2 , \quad con = \lambda ks_1. \lambda ks_2. \lambda x. ((ks_1 \ x),(ks_2 \ x))$

@@

Apart from the last construction, the others are quite similar to the corresponding ones of $pg$, but have some updates over the source and the view on accumulative functions $ks$ and $kv$ respectively.\\
For $\cpginline{bx_1 \circ bx_2}$, there are only two $cpg$ calls. The first call $\cpginline{bx_1}$ requires parameter $(ks, id, s, construct\_dummy \ s)$ where $s$ and $ks$ are corresponding to the source and the update over source. Because there is no real view here, we need to construct a dummy from the source. Then the continuation updating on this dummy should be initiated as an identity function. The first $cpg$ call is assigned to a 4-tuple $(ks_1, kv_1, s_1, v_1)$ where $s_1$ is redundant because this evaluated variable is not used in the later steps. In the next assigment, a 4-tuple $(ks_2, kv_2, s_2, v_2)$ is assigned by the second $cpg$ call which uses the input as $(kv_1, kv, v_1, v)$ where $kv_1$ and $v_1$ are obtained from the result of the first assignment, and $kv$ and $v$ come from the input. It is relatively similar to the second $pg$ call assignment in $\pginline{bx_1 \circ bx_2}$. After two $cpg$ calls, a function application, $ks_1 \ s_2$, is used to produce the updated source instead of calling recursively one more time like in $\pginline{bx_1 \circ bx_2}$.\\
Suppose that we have a source $s_0$ and a view $v_0$. The pair of the updated source and view $(s, v)$ where $s = \putbx{bx}{s_0}{v_0}$ and $v = \getbx{bx}{s_0}$ can be obtained using $cpg$ as follows:\\
    $\tab s \Leftarrow s_0; v \Leftarrow v_0; (ks, kv, s, v) \Leftarrow \cpg{bx}{\lambda \_.s}{id}{s}{v};\\
        \qtab (s; v)$\\
In general, the begining of a continuation should be an identity function. However, to be able to use the function application to get the result of $\cpginline{bx_1 \circ bx_2}$, the accumulative function on the source $s$ needs to be initiated as a constant function $\lambda \_.s$.\\
Suppose the begining of continuations $ks$ and $kv$ are $ks_0$ and $id$ respectively. Let's consider $\cpg{phead_1 \circ phead_2}{ks_0}{id}{s}{v}$ where $s = [[1,2,3], [\,], [4,5]]$ and $v = 100$. After the first two assignments in the definition of $cpg$ for the composition, we have: $ks_1 = f_2 \circ (con \ (fst \circ g_1 \circ id) \ (snd \circ f_1 \circ ks_0))$ and $s_2 = [100,2,3]$ where $f_1 = \lambda (s::ss).(s,ss)$, $f_2 = \lambda (s,ss).(s::ss)$, $g_1 = \lambda v.(v,())$. \\
Then: $ks_1 \ s_2 = (f_2 \circ (con \ (fst \circ g_1 \circ id) \ (snd \circ f_1 \circ ks_0))) \ s_2 \\
\tab = f_2 \ (\ (fst(g_1(id(s_2))) \ , \ snd(f_1(ks_0(s_2))))\ )\\
\tab = fst(g_1(id(s_2))) :: snd(f_1(ks_0(s_2))) = [100,2,3] :: snd(f_1(ks_0([100,2,3])))$\\
If $ks_0$ is an identity function, $ks_1 \ s_2 = [100,2,3] :: [2,3]$. This is an unexpected result when we target it to be the updated source. If $ks_0 = \lambda \_.s$ where $s = [[1,2,3], [\,], [4,5]]$, the result will be what we want to see.$ks_1 \ s_2 = [100,2,3] :: [[\,], [4,5]] = [[100,2,3], [\,], [4,5]]$. Through the above example, using the continuation $ks$ as a constant function at the beginning contributes to retain the discarded things in the source.

@@

The problem for $cpg$ lies in redundant computation during evaluating. For keeping away such computations, we introduce an extention named $kpg$. While $cpg$ evaluate values eagerly, $kpg$ does the opposite. Every values are evaluated lazily in a computation of $kpg$. The input of $\kpginline{bx}$ is expanded to a 6-tuple $(ks, kv, ks', kv', s, v)$ where $ks$ and $kv$ keep the modifcation information, $s$ and $v$ hold evaluated values, and $ks'$ and $kv'$ are used for lazy evaluation of actual values. The output of this function is also a 6-tuple $(ks, kv, ks', kv', s, v)$.\\
Suppose that we have a source $s_0$ and a view $v_0$. The pair of the updated source and view $(s, v)$ where $s = \putbx{bx}{s_0}{v_0}$ and $v = \getbx{bx}{s_0}$ can be obtained using $kpg$ as follows:\\
    $\tab s \Leftarrow s_0; v \Leftarrow v_0; (ks, kv, ks', kv', s, v) \Leftarrow \kpg{bx}{\lambda \_.s}{id}{id}{id}{s}{v};\\
        \qtab (ks' \ s; kv' \ v)$\\
The begining of accumulative functions $ks'$ and $kv'$ are set as identity, while $ks$ and $kv$ are initiated as the same with the corresponding ones in $cpg$.

@@

In the construction of $kpg$, $s$ and $v$ hold actual values only in case of $Skip$ and $Case$. Except them, functions for the computation will be kept in $ks'$ and $kv'$. When evaluating $\kpginline{bx_1 \circ bx_2}$, $ks_1'$ and $s_1$ in the result of the first assignment are still redundant but they are not evaluated. In $\kpginline{bx_1 \times bx_2}$, the evaluation of $ks'$ and $kv'$ will be done indendently in two assingments using $\kpginline{bx_1}$ and $\kpginline{bx_2}$. There may be some recomputations in $fst \circ ks'$ and $snd \circ ks'$ as well as $fst \circ kv'$ and $snd \circ kv'$. It is possible to evaluate actual values in $s$ and $v$ before calling $\kpginline{bx_1}$ to remove the redundancy like that.

\subsection{xpg}
The most imporant thing when using $kpg$ to evaluate a composition $bx_1 \circ bx_2$ is keeping the dropped parts from the source in a function that is used to produce new source instead of using more recursive call. This is reflected in the first call $\kpginline{bx_1}$ and the last application in $\kpginline{bx_1 \circ bx_2}$. The second call $\kpginline{bx_2}$ just play an syntactic role while it requires many parameters. We can completely reduce this number to 2, including one for the source and one for the view as in a $pg$ call. This idea is realized by function $xpg$ as follows:

@@

$\xpg{bx}{s}{v} = \text{same with } pg \text{ if } bx \neq bx_1 \circ bx_2$

$\xpg{bx_1 \circ bx_2}{s}{v} =\\
    \tab (ks_1,kv_1, ks_1', kv_1', s_1, v_1) \Leftarrow \kpg{bx_1}{\lambda \_.s}{id}{id}{id}{s}{construct\_dummy \ s};\\
    \tab (s_2, v_2) \Leftarrow \xpg{bx_2}{kv_1' \ v_1}{v};\\
    \qtab (ks_1 \ s_2, v_2)$
\end{definition}

>>>>

% $pg$ is an involution that is self-inverse. An involution is a function $f$ that satisfies $f(f(x)) = x \text{ for all } x \text{ in the domain of } f$.

@@

$\tab \pg{Skip \ h}{s}{v} \overset{1}{=} (\sif{h \ s = v}{s}{\text{fail}}, h \ s) \\
    \qtab \overset{2}{=} \sif{h \ s = v}{(s, h \ s)}{\text{fail}} \\
    \qtab \overset{3}{=} \sif{h \ s = v}{(s, v)}{\text{fail}}$\\
The first equality is simply based on the definitions of $pg$, $\putbxinline{Skip \ h}$ and $\getbxinline{Skip \ h}$. The second one tuples two results of $put$ and $get$ in the body of the if-expression. This is a trick since in some cases, the result of $pg$ may be fail although there is no fail when evaluating $\getbxinline{Skip \ h}$. The last equality is relatively oblivious.

@@

A restriction for $\pginline{Case}$ needs to be introduced here. We know that there is one entering condition and one exit condition when evaluating $\putbxinline{Case}$ as well as $\getbxinline{Case}$. If a tupling occurs, there will be 4 combinations from these conditions. This means two entering conditions of $\putbxinline{Case}$ and $\getbxinline{Case}$ are not always simultaneously satisfied. The evaluated branches are distinct in the $put$ and $get$ directions for combinations $((cond_{sv} \ s \ v) \ \&\& \ (not (cond_{s} \ s)))$ and $((not (cond_{sv} \ s \ v) \ \&\& \ (cond_{s} \ s))$, which are restricted in this paper. This does not happen for the others which is used in the construction of $\pginline{Case}$.

@@

The construction of $\pginline{bx_1 \circ bx_2}$ can be considered as the soul of the $pg$ function. The first two equalities comes from mentioned definitions and some basic transformations. The third one rewrites $v_1 \Leftarrow \getbx{bx_1}{s}$ into $(s_1, v_1) \Leftarrow \pg{bx_1}{s}{dummy}$. It is possible if we consider $\getbx{bx_1}{s}$ as the second element of $\pg{bx_1}{s}{dummy}$ where $dummy$ is a special value that makes the $put \, [\![bx_1]\!]$ valid. We can construct a $dummy$ from the source by \textsc{GetPut} law in some cases or require a valid view from programmers. The last equality changes $dummy$ by an application $construct\_dummy \ s$, and also substitutes $s$ in $(s_3, v_3) \Leftarrow \pg{bx_1}{s}{s_2}$ by $s_1$ which equals to $\putbx{bx_1}{s}{(construct\_dummy \ s)}$. This transformation is possible because of the \textsc{PutPut} law. Then, $v_3$ in the result pair $(s_3,v_3)$ equals $dummy$. So we can realize that there is no loss information when computing a composition.

@@

When evaluating $\pginline{bx_1 \circ bx_2}$, there are three $pg$ calls, of which twice for $\pginline{bx_1}$ and once for $\pginline{bx_2}$. If a given program is a left-associative composition, the number of $pg$ calls will be exponential. Therefore, the runtime inefficiency is inevitable. To solve that, we introduce a new function, $cpg$, accumulates changes in the source and the view. $\cpg{bx}{ks}{kv}{s}{v}$ is an extension of $\pg{bx}{s}{v}$ where $ks$ and $kv$ are continuations used to hold the modification information, and $s$ and $v$ are used to keep evaluated values. The output of this function is a 4-tuple $(ks, kv, s, v)$. To be more convenient for presenting the definition of $cpg$ as well as the other functions later, we provide some following utility functions:\\ $fst = \lambda (x_1,x_2). x_1 , \ snd = \lambda (x_1,x_2). x_2 , \ con = \lambda ks_1. \lambda ks_2. \lambda x. ((ks_1 \ x),(ks_2 \ x))$

@@

Apart from the last construction, the others are quite similar to the corresponding ones of $pg$, but have some updates over the source and the view on accumulative functions $ks$ and $kv$ respectively.\\
For $\cpginline{bx_1 \circ bx_2}$, there are only two $cpg$ calls. The first call $\cpginline{bx_1}$ requires parameter $(ks, id, s, construct\_dummy \ s)$ where $s$ and $ks$ are corresponding to the source and the update over source. Because there is no real view here, we need to construct a dummy from the source. Then the continuation updating on this dummy should be initiated as an identity function. The first $cpg$ call is assigned to a 4-tuple $(ks_1, kv_1, s_1, v_1)$ where $s_1$ is redundant because this evaluated variable is not used in the later steps. In the next assignment, a 4-tuple $(ks_2, kv_2, s_2, v_2)$ is assigned by the second $cpg$ call which uses the input as $(kv_1, kv, v_1, v)$ where $kv_1$ and $v_1$ are obtained from the result of the first assignment, and $kv$ and $v$ come from the input. It is relatively similar to the second $pg$ call assignment in $\pginline{bx_1 \circ bx_2}$. After two $cpg$ calls, a function application, $ks_1 \ s_2$, is used to produce the updated source instead of calling recursively one more time like in $\pginline{bx_1 \circ bx_2}$.\\
Suppose that we have a source $s_0$ and a view $v_0$. The pair of the updated source and view $(s, v)$ where $s = \putbx{bx}{s_0}{v_0}$ and $v = \getbx{bx}{s_0}$ can be obtained using $cpg$ as follows:\\
    $\tab s \Leftarrow s_0; v \Leftarrow v_0; (ks, kv, s, v) \Leftarrow \cpg{bx}{\lambda \_.s}{id}{s}{v};\\
        \qtab (s; v)$\\
In general, the beginning of a continuation should be an identity function. However, to be able to use the function application to get the result of $\cpginline{bx_1 \circ bx_2}$, the accumulative function on the source $s$ needs to be initiated as a constant function $\lambda \_.s$.

Suppose the beginning of continuations $ks$ and $kv$ are $ks_0$ and $id$ respectively. Let's consider $\cpg{phead_1 \circ phead_2}{ks_0}{id}{s}{v}$ where $s = [[1,2,3], [\,], [4,5]]$ and $v = 100$. After the first two assignments in the definition of $cpg$ for the composition, we have: $ks_1 = f_2 \circ (con \ (fst \circ g_1 \circ id) \ (snd \circ f_1 \circ ks_0))$ and $s_2 = [100,2,3]$ where $f_1 = \lambda (s::ss).(s,ss)$, $f_2 = \lambda (s,ss).(s::ss)$, $g_1 = \lambda v.(v,())$. \\
Then: $ks_1 \ s_2 = (f_2 \circ (con \ (fst \circ g_1 \circ id) \ (snd \circ f_1 \circ ks_0))) \ s_2 \\
\tab = f_2 \ (\ (fst(g_1(id(s_2))) \ , \ snd(f_1(ks_0(s_2))))\ )\\
\tab = fst(g_1(id(s_2))) :: snd(f_1(ks_0(s_2))) = [100,2,3] :: snd(f_1(ks_0([100,2,3])))$\\
If $ks_0$ is an identity function, $ks_1 \ s_2 = [100,2,3] :: [2,3]$. This is an unexpected result when we target it to be the updated source. If $ks_0 = \lambda \_.s$ where $s = [[1,2,3], [\,], [4,5]]$, the result will be what we want to see.$ks_1 \ s_2 = [100,2,3] :: [[\,], [4,5]] = [[100,2,3], [\,], [4,5]]$. Through the above example, using the continuation $ks$ as a constant function at the beginning contributes to retain the discarded things in the source.

@@

The problem for $cpg$ lies in redundant computation during evaluating. For keeping away such computations, we introduce an extension named $kpg$. While $cpg$ evaluate values eagerly, $kpg$ does the opposite. Every values are evaluated lazily in a computation of $kpg$. The input of $\kpginline{bx}$ is expanded to a 6-tuple $(ks, kv, ks', kv', s, v)$ where $ks$ and $kv$ keep the modification information, $s$ and $v$ hold evaluated values, and $ks'$ and $kv'$ are used for lazy evaluation of actual values. The output of this function is also a 6-tuple $(ks, kv, ks', kv', s, v)$.\\
Suppose that we have a source $s_0$ and a view $v_0$. The pair of the updated source and view $(s, v)$ where $s = \putbx{bx}{s_0}{v_0}$ and $v = \getbx{bx}{s_0}$ can be obtained using $kpg$ as follows:\\
    $\tab s \Leftarrow s_0; v \Leftarrow v_0; (ks, kv, ks', kv', s, v) \Leftarrow \kpg{bx}{\lambda \_.s}{id}{id}{id}{s}{v};\\
        \qtab (ks' \ s; kv' \ v)$\\
The beginning of accumulative functions $ks'$ and $kv'$ are set as identity, while $ks$ and $kv$ are initiated as the same with the corresponding ones in $cpg$.

@@

Except $\kpginline{Skip}$ and $\kpginline{Case}$ where $s$ and $v$ hold actual values, functions for the computation will be kept in $ks'$ and $kv'$. When evaluating $\kpginline{bx_1 \circ bx_2}$, $ks_1'$ and $s_1$ in the result of the first assignment are still redundant but they are not evaluated. In $\kpginline{bx_1 \times bx_2}$, the evaluation of $ks'$ and $kv'$ will be done independently in two assignments using $\kpginline{bx_1}$ and $\kpginline{bx_2}$. There may be same computations in $fst \circ ks'$ and $snd \circ ks'$ as well as $fst \circ kv'$ and $snd \circ kv'$. It is possible to evaluate actual values in $s$ and $v$ before calling $\kpginline{bx_1}$ to remove the redundancy like that. One more thing, in $\kpginline{Case}$, we need to evaluate $ks' \ s'$ and $kv' \ v'$ to check the fi condition before returning the 6-tuple. Such evaluations can be done lazily to make programs run faster. We use the above small optimizations in our implementations.

\subsection{xpg}
The most important thing when using $kpg$ to evaluate a composition $bx_1 \circ bx_2$ is keeping the dropped parts from the source in a function that is used to produce new source instead of using more recursive call. This is reflected in the first call $\kpginline{bx_1}$ and the last application in $\kpginline{bx_1 \circ bx_2}$. The second call $\kpginline{bx_2}$ just play an syntactic role while it requires many parameters. We can completely reduce this number to 2, including one for the source and one for the view as in a $pg$ call. This idea is realized by function $xpg$ as follows:

@@

$\xpg{bx}{s}{v} = \text{match } bx \text{ with }\\
    \tab \vert \ bx_1 \circ bx_2 \to \\
        \tabs{2} (ks_1,kv_1, ks_1', kv_1', s_1, v_1) \Leftarrow \kpg{bx_1}{\lambda \_.s}{id}{id}{id}{s}{construct\_dummy \ s};\\
        \tabs{2} (s_2, v_2) \Leftarrow \xpg{bx_2}{kv_1' \ v_1}{v};\\
        \tabs{3} (ks_1 \ s_2, v_2)\\
    \tab \vert \ \_ \to \text{similar to } put$
\end{definition}
